---
title: "DATS 6101 Midterm Project"
author: "Venkatesh Nagarjuna, Diluni Peiris, Manoj Kodihalli Venkatesh, Jose Luis Garcia Montero, Ella Brown"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---



```{r init, include=FALSE}
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 5) 
```

Loading Libraries

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(corrplot)
```

Reading the Data

```{r}
#fitnessdf <- data.frame(read.csv("C:\\Users\\dilun\\OneDrive\\Documents\\GWU\\Semester 1\\DATS 6101 - Introduction to Data Science\\Mid Term Project\\workout_fitness_tracker_data.csv"))
```

```{r}
fitnessdf <- data.frame(read.csv("workout_fitness_tracker_data.csv"))
```

**Data Pre-processing** \* Removing Redundant Columns \* Renaming Columns \* Updating
Data Types \* Imputation \* Handling Outliers \* Feature Creation


**Removing Redundant Columns** The following columns will be removed from the
dataset. User ID: Unique entries Water Intake (liters), VO2 Max and Body Fat (%):
Redundant data (same value repeated)

```{r}
fitnessdf_cleaned <- fitnessdf[, !names(fitnessdf) %in% c("User.ID", "Water.Intake..liters.", "VO2.Max", "Body.Fat....")] 
```

```{r}
str(fitnessdf_cleaned)
```

**Renaming Columns**

```{r}
colnames(fitnessdf_cleaned)<- c("Age", "Gender","Height_cm","Weight_kg","Workout_Type","Workout_Duration_mins","Calories_Burned",                   "Heart_Rate_bpm","Steps_Taken","Distance_km","Workout_Intensity","Sleep_Hours","Daily_Calories_Intake",      "Resting_Heart_Rate_bpm","Mood_Before_Workout","Mood_After_Workout")
```

**Updating Data Types**

```{r}
factor_cols <- c("Gender", 
                 "Workout_Type", 
                 "Workout_Intensity", 
                 "Mood_Before_Workout", 
                 "Mood_After_Workout")

fitnessdf_cleaned[factor_cols] <- lapply(fitnessdf_cleaned[factor_cols], function(x) as.factor(x))
str(fitnessdf_cleaned)
```

**Imputation**

```{r}
missing_report <- data.frame(
  Column = names(fitnessdf_cleaned),
  Missing_Count = colSums(is.na(fitnessdf_cleaned)),
  Missing_Percentage = colSums(is.na(fitnessdf_cleaned)) / nrow(fitnessdf_cleaned) * 100
  )

print(missing_report) 
```

\*There are no missing values in the dataset.

**Handling Outliers** 

```{r}
# Select numeric columns
fitnessdf_numeric <- fitnessdf_cleaned %>% select(where(is.numeric))

# Convert to long format
fitnessdf_long <- fitnessdf_numeric %>% pivot_longer(
  cols = everything(),
  names_to = "Variable",
  values_to = "Value"
)

# Create box plots in a grid layout
ggplot(fitnessdf_long, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 3) +
  theme_minimal() +
  labs(title = "Boxplots of Numeric Columns") +
  theme(legend.position = "none") +
  facet_wrap(~Variable, scales = "free",ncol= 3)  # Creates a grid of individual plots
```

Detecting outliers using Z-score.

```{r}
# Detecting outliers by Z-score
zscore_outliers <- function(df, threshold = 3) {
  outliers <- data.frame(Row = integer(), Column = character(), Value = numeric())
  
  for (col in colnames(df)) {
    if (is.numeric(df[[col]])) {
      z_scores <- scale(df[[col]])
      rows <- which(abs(z_scores) > threshold)
      
      if (length(rows) > 0) {
        for (r in rows) {
          outliers <- rbind(outliers, data.frame(Row = r, Column = col, Value = df[r, col]))
        }
      }
    }
  }
  
  return(outliers)
}

# Detect outliers using Z-score
outliers_z <- zscore_outliers(fitnessdf_cleaned)
print("Outliers detected using Z-score:")
print(outliers_z)

```

* No outliers were observed in the data.

**Feature Creation** BMI: To categorize the dataset based on BMI categroies as
Underweight, Normal, Overweight and Obese BMI Category: Four classes as Underweight,
Normal, Overweight and Obese Age Groups: Based on the dataset(Ages 18-59), 18-30
years (Youth), 31-45 (Middle age), 46-59 (Senior)

**BMI**

```{r}
fitnessdf_cleaned$BMI <- fitnessdf_cleaned$Weight_kg / (fitnessdf_cleaned$Height_cm / 100)^2
```

**BMI Category**

Segmenting the data based on BMI

```{r}

fitnessdf_cleaned <- fitnessdf_cleaned %>%
  mutate(BMI_category= case_when(
     BMI < 18.5 ~ "Underweight",
     BMI >= 18.5 & BMI < 24.9 ~ "Normal",
     BMI >= 24.9 & BMI < 29.9 ~ "Overweight",
     BMI >= 29.9 ~ "Obese"

  )) 
```


**Age Groups**

Age Groups: Based on the dataset(Ages 18-59), 18-30 years (Youth), 31-45 (Middle
age), 46-59 (Senior)

```{r, include=T}
fitnessdf_cleaned <- fitnessdf_cleaned %>%
  mutate(Age_group= case_when(
     Age <= 30 ~ "Youth",
     Age >= 31 & Age <= 45 ~ "Middle age",
     Age >= 46 ~ "Senior"
  )) 
```

```{r, include=T}
head(fitnessdf_cleaned)
```

##EDA

### EDA: Demographic Analysis

```{r}
# Summary statistics for demographic variables
demographics_summary <- fitnessdf_cleaned %>% 
  select(Age, Gender, Height_cm, Weight_kg, BMI_category) %>% 
  summary()
print(demographics_summary)

# Age Distribution Histogram
ggplot(fitnessdf_cleaned, aes(x = Age)) + 
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()
```


```{r}
# Age Category distribution 
ggplot(fitnessdf_cleaned, aes(x = Age_group, fill = Age_group)) +
  geom_bar() +
  labs(title = "Distribution of Age Groups", x = "Age_group", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")  # Adds nice colors
```


```{r}
#BMI- Category Distibution
ggplot(fitnessdf_cleaned, aes(x = BMI_category, fill = BMI_category)) +
  geom_bar() +
  labs(title = "Distribution of BMI Categories", x = "BMI Category", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")  # Adds nice colors

```


```{r}
# Gender Distribution Bar Plot (if the 'gender' column exists)
if("Gender" %in% colnames(fitnessdf_cleaned)){
  ggplot(fitnessdf_cleaned, aes(x = Gender)) +
    geom_bar(fill = "coral", color = "black", alpha = 0.7) +
    labs(title = "Gender Distribution", x = "Gender", y = "Count") +
    theme_minimal()
}
```

```{r}
# Gender Distribution based on BMI Category
ggplot(fitnessdf_cleaned,aes(x=Gender, fill= BMI_category))+
  geom_bar(position = "dodge")+
  labs(title = "Gender Distribution based on BMI Category", x = "Gender", y = "Count")+
  theme_minimal()+
  scale_fill_brewer(palette = "Set2")
  
```


### EDA: Performance Metrics

```{r}
# Summary statistics for performance metrics
performance_summary <- fitnessdf_cleaned %>% 
  select(Workout_Duration_mins, Calories_Burned, Heart_Rate_bpm) %>% 
  summary()
print(performance_summary)

# Scatter Plot: Workout Duration vs. Calorie Burn
if(all(c("Workout_Duration_mins", "Calories_Burned") %in% colnames(fitnessdf_cleaned))){
  ggplot(fitnessdf_cleaned, aes(x = Workout_Duration_mins, y = Calories_Burned)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    labs(title = "Workout Duration vs. Calorie Burn", 
         x = "Workout Duration (min)", y = "Calories Burned") +
    theme_minimal()
}
```

```{r}
# Heart Rate Distribution Histogram
if("Heart_Rate_bpm" %in% colnames(fitnessdf_cleaned)){
  ggplot(fitnessdf_cleaned, aes(x = Heart_Rate_bpm)) +
    geom_histogram(bins = 30, fill = "purple", color = "black", alpha = 0.7) +
    labs(title = "Heart Rate Distribution", 
         x = "Heart Rate (bpm)", y = "Frequency") +
    theme_minimal()
}
```

```{r}
#Boxplot: Mood before workout and sleep hours 
ggplot(fitnessdf_cleaned, aes(y = Sleep_Hours, x= Mood_Before_Workout)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "#90EE90")+
  labs(title = "Boxplot of Sleep Hours across Different Moods Before Workout", x = "Mood_Before_Workout", y = "Sleep_Hours") +
  theme_minimal() 
```

If you have had more hours of sleep could assume that your mood before workout would
be better.


### EDA: Workout Patterns

```{r}
# Frequency of Workout Types (if the 'workout_type' column exists)
if("Workout_Type" %in% colnames(fitnessdf_cleaned)){
  ggplot(fitnessdf_cleaned, aes(x = Workout_Type)) +
    geom_bar(fill = "orange", color = "black", alpha = 0.7) +
    labs(title = "Frequency of Workout Types", 
         x = "Workout Type", y = "Count") +
    theme_minimal()
  
  # Boxplot: Workout Duration by Workout Type
  ggplot(fitnessdf_cleaned, aes(x = Workout_Type, y = Workout_Duration_mins)) +
    geom_boxplot(fill = "lightblue", alpha = 0.7) +
    labs(title = "Workout Duration by Workout Type", 
         x = "Workout Type", y = "Duration (min)") +
    theme_minimal()
}
```


## EDA: Gender Mean differences

```{r}
fitnessdf_gender <- fitnessdf_cleaned %>%
  filter(Gender %in% c("Male", "Female"))

gender_summaries <- fitnessdf_gender %>%
  group_by(Gender) %>%
  summarise(
    count = n(),
    mean_height = mean(Height_cm),
    sd_height = sd(Height_cm),
    mean_weight = mean(Weight_kg),
    sd_weight = sd(Weight_kg),
    mean_calories = mean(Calories_Burned),
    sd_calories = sd(Calories_Burned),
    mean_duration = mean(Workout_Duration_mins),
    sd_duration = sd(Workout_Duration_mins)
  )

print("Summary Statistics by Gender:")
print(gender_summaries)

# Perform t-tests for various metrics
t_test_results <- list(
  Height = t.test(Height_cm ~ Gender, data = fitnessdf_gender),
  Weight = t.test(Weight_kg ~ Gender, data = fitnessdf_gender),
  Duration = t.test(Workout_Duration_mins ~ Gender, data = fitnessdf_gender),
  Calories = t.test(Calories_Burned ~ Gender, data = fitnessdf_gender),
  Heart_Rate = t.test(Heart_Rate_bpm ~ Gender, data = fitnessdf_gender),
  Steps = t.test(Steps_Taken ~ Gender, data = fitnessdf_gender),
  Distance = t.test(Distance_km ~ Gender, data = fitnessdf_gender),
  Sleep = t.test(Sleep_Hours ~ Gender, data = fitnessdf_gender),
  Daily_Calories = t.test(Daily_Calories_Intake ~ Gender, data = fitnessdf_gender),
  Resting_HR = t.test(Resting_Heart_Rate_bpm ~ Gender, data = fitnessdf_gender)
)

# Create a summary table of t-test results
t_test_summary <- data.frame(
  Variable = names(t_test_results),
  P_Value = sapply(t_test_results, function(x) x$p.value),
  Mean_Difference = sapply(t_test_results, function(x) diff(x$estimate)),
  T_Statistic = sapply(t_test_results, function(x) x$statistic)
)

# Add significance indicators
t_test_summary$Significance <- ifelse(t_test_summary$P_Value < 0.001, "***",
                                    ifelse(t_test_summary$P_Value < 0.01, "**",
                                           ifelse(t_test_summary$P_Value < 0.05, "*", "ns")))

print("T-Test Results Summary:")
print(t_test_summary)
    
```

##Statistical Testing

**SMART Q1, Q2, Q3**

**SMART Q4**

1\. Effects of workout type on different BMI categories [Factorial ANOVA]

Factor 1: BMI Category - Underweight, Normal, Overweight and Obese

Factor 2 : Workout Type - Cycling, Cardio, HIIT, Strength, Yoga, Running

Dependent Variable: Calories_Burned

[Main Effects]{.underline}

BMI Category:

$$
H_0: \mu_{underweight} = \mu_{normal} = \mu_{overweight} = \mu_{obese}
$$\
$$
H_1: \text{ At least one mean is different}
$$

The calorie burn is the same for all BMI levels.

```{r}
#Boxplot: Calories_Burned across different BMI levels 
ggplot(fitnessdf_cleaned, aes(y = Calories_Burned, x= BMI_category)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "#90EE90")+
  labs(title = "Boxplot of Calories_Burned across different BMI Categories", x = "BMI_category", y = "Calories_Burned") +
  theme_minimal() 
```

Workout Type:

$$
H_0: \mu_{cycling} = \mu_{Cardio} = \mu_{HIIT} = \mu_{Strengh} = \mu_{yoga} = \mu_{running} 
$$

$$
H_1: \text{ At least one mean is different}
$$

The calorie burn is the same for all workout types.

```{r}
#Boxplot: Boxplot ofCalories_Burned across different Workout types 
ggplot(fitnessdf_cleaned, aes(y = Calories_Burned, x= Workout_Type)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "lightblue")+
  labs(title = "Boxplot of Calories_Burned across different Workout types", x = "Workout_type", y = "Calories_Burned") +
  theme_minimal() 
```

[Interaction Effect]{.underline}

$$
H_0:  \text{There is no interaction between BMI level and workout type on Calorie burn}
$$

$$
H_1:  \text{There is an interaction between BMI level and workout type on Calorie burn}
$$

The workout type does not have an influence on Calorie burn based on BMI Levels.

```{r, include=T}
# Run factorial ANOVA
result <- aov(Calories_Burned ~ BMI_category * Workout_Type, data = fitnessdf_cleaned)
```

```{r, include=T}
# Run factorial ANOVA
summary(result)
```

Based on the Results;(Testing at 95% significance level)

1\. Main effects

-   BMI Level: Cannot Reject HO as P value is \> 0.05,The average Calorie burn is
    equal among different BMI levels.
-   Workout Type: Cannot Reject HO as P value is \> 0.05, The average Calorie  burn is
    equal among different Workout types.

2.  Interaction Effect

-   Cannot Reject HO as P value is \> 0.05, Different BMI groups respond the same to
    different workout types.

**SMART Q5** Link between poor sleep and resting heart rate (fewer the hours of sleep and higher resting heart rate)
Source: <https://pmc.ncbi.nlm.nih.gov/articles/PMC7680518/>


```{r, include=T}
# Scatter Plot
ggplot(fitnessdf_cleaned, aes(x = Sleep_Hours, y = Resting_Heart_Rate_bpm)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    labs(title = "Sleep Hours vs. Resting Heart Rate (bpm)", 
         x = "Sleep_Hours", y = "Resting Heart Rate (bpm)") +
    theme_minimal()
```

```{r, include=T}
cor(fitnessdf_cleaned$Sleep_Hours,fitnessdf_cleaned$Resting_Heart_Rate_bpm)
```

Coefficient - -0.0125. It is negatively correlated as expected, however the strength
of the relationship is low.



**SMART Q6** Is there a relationship between mood before and mood after workout?

Mood Before Workout: Happy, Neutral, Stressed and Tired.

Mood After Workout: Energized, Fatigued and Neutral

Using a Chi-squared test of Independence.

$$
H_0: \text{There is no relationship between mood before and after workout} 
$$

$$
H_1: \text{There is a relationship between mood before and after workout} 
$$

```{r, include=T}
# Creating a Contingency Table
table_data <- table(fitnessdf_cleaned$Mood_Before_Workout, fitnessdf_cleaned$Mood_After_Workout)
print(table_data)
```

```{r, include=T}
# Run Chi-Square Test
chi_sq_result <- chisq.test(table_data)
print(chi_sq_result)
```

P-value is \> 0.05, Therefore we cannot reject the Null. There's no relationship
between mood before and after workout.


