---
title: "DATS 6101 Midterm Project"
author: "Venkatesh Nagarjuna, Diluni Peiris, Manoj Kodihalli Venkatesh, Jose Luis Garcia Montero,
Ella Brown"
output: html_document
runtime: shiny
editor_options: 
  markdown: 
    wrap: 85
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This R Markdown document is made interactive using Shiny. Unlike the more traditional
workflow of creating static reports, you can now create documents that allow your
readers to change the assumptions underlying your analysis and see the results
immediately.

To learn more, see [Interactive
Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Inputs and Outputs

You can embed Shiny inputs and outputs in your document. Outputs are automatically
updated whenever inputs change. This demonstrates how a standard R plot can be made
interactive by wrapping it in the Shiny `renderPlot` function. The `selectInput` and
`sliderInput` functions create the input widgets used to drive the plot.

```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document
using the `shinyAppDir` function. This example embeds a Shiny application located in
another directory:

```{r tabsets, echo=FALSE}
shinyAppDir(
  system.file("examples/06_tabsets", package = "shiny"),
  options = list(
    width = "100%", height = 550
  )
)
```

Note the use of the `height` parameter to determine how much vertical space the
embedded application should occupy.

You can also use the `shinyApp` function to define an application inline rather then
in an external directory.

In all of R code chunks above the `echo = FALSE` attribute is used. This is to
prevent the R code within the chunk from rendering in the document alongside the
Shiny components.

Loading Libraries

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
```

Reading the Data

```{r}
fitnessdf <- data.frame(read.csv("workout_fitness_tracker_data.csv"))
```

**Data Pre-processing** \* Removing Redundant Columns \* Renaming Columns \* Updating
Data Types \* Imputation \* Handling Outliers \* Variable Scaling \* Feature Creation
\* Feature Selection

**Removing Redundant Columns** The following columns will be removed from the
dataset. User ID: Unique entries Water Intake (liters), VO2 Max and Body Fat (%):
Redundant data (same value repeated)

```{r}
fitnessdf_cleaned <- fitnessdf[, !names(fitnessdf) %in% c("User.ID", "Water.Intake..liters.", "VO2.Max", "Body.Fat....")] 
```

```{r}
str(fitnessdf_cleaned)
```

**Renaming Columns**

```{r}
colnames(fitnessdf_cleaned)<- c("Age", "Gender","Height_cm","Weight_kg","Workout_Type","Workout_Duration_mins","Calories_Burned",
                                "Heart_Rate_bpm","Steps_Taken","Distance_km","Workout_Intensity","Sleep_Hours","Daily_Calories_Intake",
                                "Resting_Heart_Rate_bpm","Mood_Before_Workout","Mood_After_Workout")
```

**Updating Data Types**

```{r}
factor_cols <- c("Gender", 
                 "Workout_Type", 
                 "Workout_Intensity", 
                 "Mood_Before_Workout", 
                 "Mood_After_Workout")

fitnessdf_cleaned[factor_cols] <- lapply(fitnessdf_cleaned[factor_cols], function(x) as.factor(x))
str(fitnessdf_cleaned)
```

**Imputation**

```{r}
missing_report <- data.frame(
  Column = names(fitnessdf_cleaned),
  Missing_Count = colSums(is.na(fitnessdf_cleaned)),
  Missing_Percentage = colSums(is.na(fitnessdf_cleaned)) / nrow(fitnessdf_cleaned) * 100
  )

print(missing_report) 
```

\*There are no missing values in the dataset.

**Handling Outliers** TBD

```{r}
# Select numeric columns
fitnessdf_numeric <- fitnessdf_cleaned %>% select(where(is.numeric))

# Convert to long format
fitnessdf_long <- fitnessdf_numeric %>% pivot_longer(
  cols = everything(),
  names_to = "Variable",
  values_to = "Value"
)

# Create box plots in a grid layout
ggplot(fitnessdf_long, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 3) +
  theme_minimal() +
  labs(title = "Boxplots of Numeric Columns") +
  theme(legend.position = "none") +
  facet_wrap(~Variable, scales = "free",ncol= 3)  # Creates a grid of individual plots
```

**Variable Scaling** TBD

**Feature Creation** BMI: To categorize the dataset based on BMI categroies as
Underweight, Normal, Overweight and Obese BMI Category: Four classes as Underweight,
Normal, Overweight and Obese Age Groups: Based on the dataset(Ages 18-59), 18-30
years (Youth), 31-45 (Middle age), 46-59 (Senior)

**BMI**

```{r}
fitnessdf_cleaned$BMI <- fitnessdf_cleaned$Weight_kg / (fitnessdf_cleaned$Height_cm / 100)^2
```

**BMI Category**

Segmentaing the data based on BMI

```{r}

fitnessdf_cleaned <- fitnessdf_cleaned %>%
  mutate(BMI_category= case_when(
     BMI < 18.5 ~ "Underweight",
     BMI >= 18.5 & BMI < 24.9 ~ "Normal",
     BMI >= 24.9 & BMI < 29.9 ~ "Overweight",
     BMI >= 29.9 ~ "Obese"

  )) 
```

Bar Plot

```{r}

ggplot(fitnessdf_cleaned, aes(x = BMI_category, fill = BMI_category)) +
  geom_bar() +
  labs(title = "Distribution of BMI Categories", x = "BMI Category", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")  # Adds nice colors

```

**Age Groups**

Age Groups: Based on the dataset(Ages 18-59), 18-30 years (Youth), 31-45 (Middle
age), 46-59 (Senior)

```{r Q9, include=T}
fitnessdf_cleaned <- fitnessdf_cleaned %>%
  mutate(Age_group= case_when(
     Age <= 30 ~ "Youth",
     Age >= 31 & Age <= 45 ~ "Middle age",
     Age >= 46 ~ "Senior"
  )) 
```

```{r Q9, include=T}
head(fitnessdf_cleaned)
```

```{r Q9, include=T}
ggplot(fitnessdf_cleaned, aes(x = Calories_Burned)) +
  geom_histogram(binwidth = 50, color = "black", fill = "green", alpha = 0.7) +
  labs(title = "Histogram of Baseball Player Weights", x = "Weight", y = "Frequency") +
  theme_minimal()
```

```{r}

#Creating a boxplot for gre Scores of Admitted Students Across School Rankings
ggplot(fitnessdf_cleaned, aes(y = Calories_Burned, x= Workout_type)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "#90EE90")+
  labs(title = "Boxplot of gre Scores of Admitted Students Across School Rankings", x = "School Rank", y = "gre Score") +
  theme_minimal() 
```

##EDA


### EDA: Demographic Analysis

```{r}
# Summary statistics for demographic variables
demographics_summary <- fitnessdf_cleaned %>% 
  select(Age, Gender, Height_cm, Weight_kg, BMI_category) %>% 
  summary()
print(demographics_summary)

# Age Distribution Histogram
ggplot(fitnessdf_cleaned, aes(x = Age)) + 
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()

# Gender Distribution Bar Plot (if the 'gender' column exists)
if("gender" %in% colnames(fitnessdf_cleaned)){
  ggplot(fitnessdf_cleaned, aes(x = Gender)) +
    geom_bar(fill = "coral", color = "black", alpha = 0.7) +
    labs(title = "Gender Distribution", x = "Gender", y = "Count") +
    theme_minimal()
}
```


### EDA: Performance Metrics

```{r}
# Summary statistics for performance metrics
performance_summary <- fitnessdf_cleaned %>% 
  select(Workout_Duration_mins, Calories_Burned, Heart_Rate_bpm) %>% 
  summary()
print(performance_summary)

# Scatter Plot: Workout Duration vs. Calorie Burn
if(all(c("Workout_Duration_mins", "Calories_Burned") %in% colnames(fitnessdf_cleaned))){
  ggplot(fitnessdf_cleaned, aes(x = workout_duration, y = Calories_Burned)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    labs(title = "Workout Duration vs. Calorie Burn", 
         x = "Workout Duration (min)", y = "Calories Burned") +
    theme_minimal()
}

# Heart Rate Distribution Histogram
if("heart_rate" %in% colnames(fitnessdf_cleaned)){
  ggplot(fitnessdf_cleaned, aes(x = Heart_Rate_bpm)) +
    geom_histogram(bins = 30, fill = "purple", color = "black", alpha = 0.7) +
    labs(title = "Heart Rate Distribution", 
         x = "Heart Rate (bpm)", y = "Frequency") +
    theme_minimal()
}
```

### EDA: Workout Patterns

```{r}
# Frequency of Workout Types (if the 'workout_type' column exists)
if("Workout_type" %in% colnames(fitnessdf_cleaned)){
  ggplot(fitnessdf_cleaned, aes(x = Workout_type)) +
    geom_bar(fill = "orange", color = "black", alpha = 0.7) +
    labs(title = "Frequency of Workout Types", 
         x = "Workout Type", y = "Count") +
    theme_minimal()
  
  # Boxplot: Workout Duration by Workout Type
  ggplot(fitnessdf_cleaned, aes(x = Workout_type, y = Workout_Duration_mins)) +
    geom_boxplot(fill = "lightblue", alpha = 0.7) +
    labs(title = "Workout Duration by Workout Type", 
         x = "Workout Type", y = "Duration (min)") +
    theme_minimal()
}
```

##Statistical Testing

**SMART Q4**

1\. Does the workout type affects differently Relationship between workout type and
BMI Category

Effects of workout type on different BMI categories [Factorial ANOVA]

Factor 1: BMI Category - Underweight, Normal, Overweight and Obese

Factor 2 : Workout Type - Cycling, Cardio, HIIT, Strength, Yoga, Running

Dependent Variable: Calories_Burned

[Main Effects]{.underline}

BMI Category:

$$
H_0: \mu_{underweight} = \mu_{normal} = \mu_{overweight} = \mu_{obese}
$$\
$$
H_1: \text{ At least one mean is different}
$$

The calorie burn is the same for all BMI levels.

```{r}
#Boxplot: Calories_Burned across different BMI levels 
ggplot(fitnessdf_cleaned, aes(y = Calories_Burned, x= BMI_category)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "#90EE90")+
  labs(title = "Boxplot of Calories_Burned across different BMI Categories", x = "BMI_category", y = "Calories_Burned") +
  theme_minimal() 
```

Workout Type:

$$
H_0: \mu_{cycling} = \mu_{Cardio} = \mu_{HIIT} = \mu_{Strengh} = \mu_{yoga} = \mu_{running} 
$$

$$
H_1: \text{ At least one mean is different}
$$

The calorie burn is the same for all workout types.

```{r}
#Boxplot: Boxplot ofCalories_Burned across different Workout types 
ggplot(fitnessdf_cleaned, aes(y = Calories_Burned, x= Workout_type)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "lightblue")+
  labs(title = "Boxplot of Calories_Burned across different Workout types", x = "Workout_type", y = "Calories_Burned") +
  theme_minimal() 
```

[Interaction Effect]{.underline}

$$
H_0:  \text{There is no interaction between BMI level and workout type on Calorie burn}
$$

$$
H_1:  \text{There is an interaction between BMI level and workout type on Calorie burn}
$$

The workout type does not have an influence on Calorie burn based on BMI Levels.

```{r, include=T}
# Run factorial ANOVA
result <- aov(Calories_Burned ~ BMI_category * Workout_type, data = fitnessdf_cleaned)
```

```{r, include=T}
# Run factorial ANOVA
summary(result)
```

Based on the Results;(Testing at 95% significance level)

1\. Main effects

-   BMI Level: Cannot Reject HO as P value is > 0.05,The avarage Calarie burn is equal among different BMI levels.
-   Workout Type: Cannot Reject HO as P value is > 0.05, The avarage Calarie burn is equal among different Workout types.

2. Interaction Effect

-  Cannot Reject HO as P value is > 0.05, Different BMI groups respond the same to different workout types.


**SMART Q5**
Link between poor sleep (fewer hours and higher resting heart rate)
Source: https://pmc.ncbi.nlm.nih.gov/articles/PMC7680518/

Maybe remove this
```{r, include=T}
# Scatter Plot
ggplot(fitnessdf_cleaned, aes(x = Sleep_Hours, y = Resting_Heart_Rate_bpm)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    labs(title = "Sleep Hours vs. Resting Heart Rate (bpm)", 
         x = "Sleep_Hours", y = "Resting Heart Rate (bpm)") +
    theme_minimal()
```


```{r, include=T}
cor(fitnessdf_cleaned$Sleep_Hours,fitnessdf_cleaned$Resting_Heart_Rate_bpm)
```

Coefficient - -0.0125. It is negatively correlated as expected, however the strength of the relationship is low.
