---
title: "Final Project- SMART Q4"
author: "Diluni Peiris"
date: "18/4/2025"
# date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r include=FALSE}
library(car)
library(ggplot2)
library(reshape2)
library(corrplot)
library(dplyr)
library(randomForest)
library(lme4)         
library(fixest)        
library(glmnet)       
library(plotmo)  
```


# SMART Q4

## Importing Dataset 

```{r include=FALSE}
Edu <- data.frame(read.csv("SMART Q4 Data.csv"))
```


```{r}
str(Edu)
```

```{r}
ggplot(Edu, aes(x = distance)) +
  geom_bar( fill = "steelblue") +
  labs(
    title = "Distribution of Learning types",
    x = "Learning Type"
      ) +
  theme_minimal() +
  theme(axis.text.x = element_text( hjust = 1))
```

## Data Cleaning 

### 1. Distance Variable 

* Removing cataegory 0 : Field of study not reported to IPEDS
* Combining 1 and 2 : Not all can be done online in this field of study 

```{r}
df_clean <- Edu %>%
  filter(distance != 0 ) %>%
  mutate(distance = case_when(
    distance %in% c(1, 2) ~ "Most credentials onsite",
    distance == 3 ~ "All credentials online"
  ))

```


### 2. Filtering STEM fields

```{r}
df_clean <- subset(df_clean, STEM.Classification == "STEM")
```


### 3. Drop and Rename Variables

* unitid and instnm indicates the same dropping instnm.
* Dropping STEM.Classification column as it only contains STEM fields.
* Updating PS and null
* Renaming columns

```{r}
df <- subset(df_clean, select = -instnm)

df <- subset(df, select = -STEM.Classification)

# Replace all occurrences of "PS" with 0 in the entire data frame
df[df == "PS"] <- 0

#Renaming Variables
new_names <- c("school_id", "course_field","credit_level", "delivery_mode", "school_type", "earnings", "no_of_pell_grants")

colnames(df) <- new_names

```


### 4. Updating variable types
```{r}
df$school_id <- factor(df$school_id)
df$course_field <- factor(df$course_field)
df$credit_level<- factor(df$credit_level)
df$delivery_mode <- factor(df$delivery_mode,
                                        levels = c("Most credentials onsite", "All credentials online"))
df$school_type <- factor(df$school_type)
df$earnings <- as.integer(df$earnings)
df$no_of_pell_grants <- as.integer(df$no_of_pell_grants)

str(df)

```

### 5. Identifying and Remocing Outliers of the target variable [Revenue]
```{r}
# Create a boxplot to visualize the outliers
boxplot(df$earnings, main = "Boxplot of Earnings", ylab = "Earnings")
```

### Removing outliers in Earnings
```{r}
# Calculate the Z-scores for the 'earnings' column
z_scores <- (df$earnings - mean(df$earnings, na.rm = TRUE)) / sd(df$earnings, na.rm = TRUE)

# Identify outliers (values with |Z| > 3)
outliers_zscore <- df$earnings[abs(z_scores) > 3]

# Count the number of outliers
num_outliers_zscore <- length(outliers_zscore)
num_outliers_zscore

# Filter out the rows where |Z| <= 3 (i.e., remove outliers)
df <- df[abs(z_scores) <= 3, ]
```


## EDA

```{r}
ggplot(df, aes(x = delivery_mode)) +
  geom_bar( fill = "steelblue") +
  labs(
    title = "Distribution of Learning types",
    x = "Learning Type"
      ) +
  theme_minimal() +
  theme(axis.text.x = element_text( hjust = 1))
```

```{r}
ggplot(df, aes(x = credit_level)) +
  geom_bar( fill = "steelblue") +
  labs(
    title = "Distribution of credit_level",
    x = "credit_level"
      ) +
  theme_minimal() +
  theme(axis.text.x = element_text( hjust = 1))
```

```{r}
ggplot(df, aes(x = school_type)) +
  geom_bar( fill = "steelblue") +
  labs(
    title = "Distribution of school_type",
    x = "school_type"
      ) +
  theme_minimal() +
  theme(axis.text.x = element_text( hjust = 1))
```

```{r}
# Visualize earnings distribution by education type
ggplot(df, aes(x = earnings, fill = delivery_mode)) +
  geom_density(alpha = 0.5) +
  labs(title = "Earnings Distribution by Education Type",
       x = "Annual Earnings ($)",
       y = "Density") +
  theme_minimal()
```

## Model Building

### Model 1

```{r}
df2 <- subset(df, select = -school_id)
```


```{r}
#Train test split
library(caret)

set.seed(123)

# Create partition
train_index <- createDataPartition(df$earnings, p = 0.7, list = FALSE)

# Split the data
train_data <- df2[train_index, ]
test_data  <- df2[-train_index, ]

```


```{r}
model1 <- lm(earnings ~ ., data = train_data)
summary(model1)
```

### 1. Model without controls 

```{r}
model1 <- lm(earnings ~ delivery_mode, data = train_data)
summary(model1)
```
### 2. Model with field level characteristics
```{r}
model2 <- lm(earnings ~ delivery_mode + course_field, data = train_data)
summary(model2)
```


### 3. Model with field level + institutional characteristics

```{r}
model3 <- lm(earnings ~ ., data = train_data)
summary(model3)
```

### 4. Regularization 

1. Ridge Regression
```{r}
# 1. Prepare data for regularized regression

# Create model matrix for glmnet (converts factors to dummy variables)
# Remove the response variable (median_earnings) 
X <- model.matrix(earnings ~ . , data = df)[,-1]
y <- df$earnings

# Split data into training and testing sets
set.seed(123)
train_indices <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_indices, ]
y_train <- y[train_indices]
X_test <- X[-train_indices, ]
y_test <- y[-train_indices]

# Ridge Regression

# Finding optimal lambda using cross-validation
set.seed(123)
ridge_cv <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
plot(ridge_cv)

# Best lambda value
best_lambda_ridge <- ridge_cv$lambda.min
cat("Best lambda for ridge regression:", best_lambda_ridge, "\n")

# Fit ridge model with optimal lambda
ridge_model <- glmnet(X_train, y_train, alpha = 0, lambda = best_lambda_ridge)
print(ridge_model)

# Get coefficients
ridge_coefs <- coef(ridge_model)
ridge_coefs_df <- data.frame(
  feature = rownames(ridge_coefs),
  coefficient = as.vector(ridge_coefs)
) %>%
  filter(coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

print(ridge_coefs_df)

# Visualize top coefficients
ridge_coefs_df %>%
  filter(feature != "(Intercept)") %>%
  top_n(15, abs(coefficient)) %>%
  ggplot(aes(x = reorder(feature, abs(coefficient)), y = coefficient)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Ridge Regression Coefficients",
       x = "Feature",
       y = "Coefficient Value") +
  theme_minimal()

# Evaluate ridge model on test data
ridge_predictions <- predict(ridge_model, newx = X_test)
ridge_rmse <- sqrt(mean((y_test - ridge_predictions)^2))
cat("Ridge Regression RMSE:", ridge_rmse, "\n")


```
2. Lasso Regression
```{r}
# Find optimal lambda using cross-validation
set.seed(123)
lasso_cv <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
plot(lasso_cv)

# Best lambda value
best_lambda_lasso <- lasso_cv$lambda.min
cat("Best lambda for lasso regression:", best_lambda_lasso, "\n")

# Fit lasso model with optimal lambda
lasso_model <- glmnet(X_train, y_train, alpha = 1, lambda = best_lambda_lasso)
print(lasso_model)

# Get coefficients - lasso will zero out some coefficients
lasso_coefs <- coef(lasso_model)
lasso_coefs_df <- data.frame(
  feature = rownames(lasso_coefs),
  coefficient = as.vector(lasso_coefs)
) %>%
  filter(coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

print(lasso_coefs_df)

# Visualize non-zero coefficients
lasso_coefs_df %>%
  filter(feature != "(Intercept)") %>%
  ggplot(aes(x = reorder(feature, abs(coefficient)), y = coefficient)) +
  geom_col() +
  coord_flip() +
  labs(title = "Non-Zero Lasso Regression Coefficients",
       subtitle = "Variables selected by lasso as important predictors",
       x = "Feature",
       y = "Coefficient Value") +
  theme_minimal()

# Evaluate lasso model on test data
lasso_predictions <- predict(lasso_model, newx = X_test)
lasso_rmse <- sqrt(mean((y_test - lasso_predictions)^2))
cat("Lasso Regression RMSE:", lasso_rmse, "\n")
```
### Random Forest model[Results in an error]

Error in randomForest.default(m, y, ...) : 
  Can not handle categorical predictors with more than 53 categories.

```{r}
# Training a random forest model for comparison 
rf_model <- randomForest(
  earnings ~ .,
  data = df[train_indices, ],
  ntree = 500,
  importance = TRUE
)
rf_predictions <- predict(rf_model, education_data[-train_indices, ])
rf_rmse <- sqrt(mean((y_test - rf_predictions)^2))
```

### Model Performance 

1. RMSE
```{r}
model_comparison <- data.frame(
  model = c("Ridge", "Lasso", "model1", "model2", "model3"),
  rmse = c(ridge_rmse, lasso_rmse)
))
```




1. Residual Plots
```{r}
plot(model3)
```