---
title: "Final Project- Summary Paper"
author: "Jose Luis Garcia Montero, Diluni Peiris,Ella Brown,Venkatesh Nagarjuna,Manoj Kodihalli Venkatesh"
date: "26/4/2025"
# date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r include=FALSE}
library(car)
library(ggplot2)
library(reshape2)
library(corrplot)
library(dplyr)
library(randomForest)
library(lme4)         
library(fixest)        
library(glmnet)       
library(plotmo)  
```

# Team Galileo - Final Project Summary Paper

# Introduction

## Background 

The U.S. higher education landscape encompasses a wide range of credentials, from certificates to doctoral degrees, with both the choice of major and the level of credential significantly influencing median earnings outcomes. Pell Grants, which provide crucial financial support to approximately seven million low-income students each year with an average award of around \$6,000, play a vital role in access to education, but often serve as an indicator of lower post-graduation income. Additionally, the rapid rise in online program enrollment—from roughly 25% before 2020 to over 50% in 2021—has sparked important questions about the long-term earnings potential of virtual versus in-person learning pathways.

## Dataset Overview

Data sourced from U.S. Department of Education's College Scorecard Field of Study dataset. The U.S. Department of Education's College Scorecard Field of Study dataset contains over 100,000 records representing postsecondary programs across the United States. This rich dataset includes key variables such as institution name, field of study (CIP code), credential level, median earnings one year after graduation, student debt levels, Pell Grant recipient status, and program size. It offers a valuable resource for analyzing the relationships between educational choices and economic outcomes, including how factors like major, credentials, and financial aid affect post-college earnings.

Link to dataset: <https://collegescorecard.ed.gov/data/> 

### Key Variables [Add to this list of you have used any other variables in your models]

| **Variable Name** | **Description** | **Details** |
|----------------------|--------------------|-------------------------------|
| `Distance` | Distance education availability | Combined values 1 and 2 into "Some/Partial Online"; removed 0 ("Not reported") |
| `Control` | Type of institution | 1 = Public, 2 = Private Nonprofit, 3 = Private For-Profit, 4 = Foreign |
| `Credlev` | Credential level | 1–8 scale: From certificates to doctoral and professional degrees |
| `unitid` | Institution identifier | Unique ID for \~1,000 institutions |
| `cipdesc...4` | Field of study (CIP description) | Originally 385 unique fields; filtered to 152 STEM fields for some models. |
| `EARN_COUNT_PELL_WNE_5YR` | Pell Grant graduate count (working & not enrolled, 5 years post-grad) | Number of Pell recipients working and not enrolled 5 years after completion |
| `EARN_MDN_5YR` | Median earnings (working & not enrolled, 5 years post-grad) | Earnings for all graduates (not just Pell recipients) 5 years after graduation |

# EDA [ADD Graphs]

-   Distribution of earnings

    -   The median earnings (EARN_MDN) are right‐skewed, with a long tail of high‐earning programs.\
        A histogram of EARN_MDN showed most programs clustering between \$20 k and \$60 k, with a few above \$100 k ​.

-   Delivery mode comparison

    -   Boxplots reveal that fully on-site programs tend to have higher median earnings than fully online or hybrid programs, though outliers exist in each group ​.

-   Credential level trends

    -   Earnings generally rise with higher credential levels (from certificates up to doctoral degrees), suggesting a positive association between credential level and post‐grad pay ​.

-   Pell‐grant share

    -   A scatterplot of PCTPELL vs. EARN_MDN hints at a weak negative correlation, but many p-values exceeded 0.05, indicating stronger socioeconomic controls may be needed ​.

# SMART Q1 [add code]

**Can earnings be predicted based on field, degree type, and aid demographics?**

## 1. Model Selection and Determination

we tested progressively complex regression frameworks:

1.  Baseline OLS (Model 1)\
    \`earnings \~ delivery_mode\`

    -   Very low explanatory power (R² = 0.001, RMSE ≈ \$29 039) ​.

2.  Add field‐level characteristics (Model 2)\
    \`earnings \~ delivery_mode + course_field\`

    -   Improvement to R² = 0.159; Adj R² = 0.139 ​.

3.  Full OLS with institutional controls (Model 3)\
    \`earnings \~ delivery_mode + course_field + school_type + credlev + EARN_COUNT_PELL_WNE_5YR\`

    -   Best OLS performance: R² = 0.293; Adj R² = 0.275 ​.

Because of multicollinearity and the high-dimensional “course_field” factor (152 STEM fields), we also fitted:

-   Ridge regression (α tuned by cross-validation; RMSE ≈ \$24 191)

-   Lasso regression (α tuned; RMSE ≈ \$24 065; 40.7% of predictors shrunk to zero) ​

Lasso was chosen for final deployment due to its balance of predictive accuracy and feature selection.

## 2. Predictions and Examples

|  |  |  |  |  |  |
|----|----|----|----|----|----|
| Scenario | delivery_mode | school_type | credlev | Pell count | Predicted EARN_MDN |
| A | On-site | Public | Bachelor (5) | Median Pell | \$45 000 |
| B | Fully online | Private NP | Certificate (1) | High Pell | \$28 000 |
| C | Hybrid | Public | Master’s (6) | Low Pell | \$55 000 |

Using the Lasso model, predictions can be made by plugging in program attributes:Note: These values are illustrative; actual predictions follow the fitted coefficient vector.

## 3. Reliability and Model Evaluation

-   Goodness-of-fit

    -   Lasso RMSE (\~\$24 065) is \~50% of the overall median earnings, indicating moderate predictive precision.

    -   Adjusted R² of the best OLS (Model 3) is 0.275, so \~27.5% of variance is explained.

-   Statistical significance

    -   Delivery mode, credential level, and school type emerged as significant predictors (p \< 0.05).

    -   “Field of study” effects were often non-significant, suggesting unobserved heterogeneity within fields (e.g., program quality).

-   Assumption checks

    -   Residual plots showed some heteroscedasticity; normality of residuals was acceptable but warrants transformation or robust standard errors.

-   Cross-validation

    -   RMSE estimates were stable across 5-fold CV, indicating no extreme overfitting.

## 4. Recommendations for Improvement [I think we can keep this section common to all models]

1.  Additional predictors

    -   Incorporate regional labor‐market indicators (e.g., local unemployment rate, cost of living).

    -   Add institutional selectivity metrics (e.g., admit rate, graduation rate).

2.  Model enhancements

    -   Explore non-linear/ensemble methods (random forests, gradient boosting) to capture complex interactions.

    -   Test income outcomes at 3- and 5-year post-grad horizons for temporal robustness.

3.  Address limitations

    -   Control for student‐level covariates (GPA, work experience) if available.

    -   Use propensity‐score matching to reduce selection bias between online vs. on-site cohorts.

    -   Segment analyses by field clusters (e.g., STEM vs. humanities) to uncover heterogeneity.

# SMART Q2

# SMART Q3

**Do online (distance education) programs yield significantly different earnings compared to in-person programs?**

With the rapid growth in online-based education programs, we wanted to analyse if there's a significant difference between the earnings of someone who completed their education online vs onsite.

Reading the cleaned dataset

```{r include=FALSE}
df <- data.frame(read.csv("cleaned_data.csv"))
```

### [Dataset Scope]{.underline}

1.  Includes only STEM-based fields.
2.  Contains 1000 institutes across the US.

Distribution of Learning Types

```{r}
ggplot(df, aes(x = delivery_mode)) +
  geom_bar( fill = "steelblue") +
  labs(
    title = "Distribution of Learning Types",
    x = "Learning Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text( hjust = 1))
```

### [Data Structure]{.underline}

-   Target Variable: Earnings (post graduation)
-   Education Type (online vs. in-person)
-   Control variables:
    -   Field of study (STEM fields)
    -   School type (ex, Public, Private-for-profit, Private-not-for-profit, Foreign)
    -   Level of credential (ex, Undergraduate Diploma to Doctoral Degree)
    -   The number of Pell Grant recipients of that institute in that particular field.

## 1. Model Selection and Determination

1.  Model without controls

$$
Earnings \sim DeliveryMode
$$

This model was built to get an initial understanding of the impact of delivery mode on earnings, isolating it from other effects.

2.  Model with field-level characteristics

$$
Earnings \sim DeliveryMode + FieldOfStudy
$$

The earnings can vary based on the field of study, This model was built to understand the effect on delivery mode given the field of study as well.

3.  Model with field-level and institutional characteristics

$$
Earnings \sim DeliveryMode + FieldOfStudy + SchoolType + LevelOfCredential + NoOfPellGrants 
$$

The above model controls for other institutional effects that can affect earnings.

4.  Models with Regularization

    4.1 Ridge Regression - To stabilize estimates, as there are many field and institutional level factors.

    4.2 Lasso Regression - To perform feature selection

## 2. Predictions and Examples

```{r}
#Train test split
library(caret)

set.seed(123)

# Create partition
train_index <- createDataPartition(df$earnings, p = 0.7, list = FALSE)

# Split the data
train_data <- df[train_index, ]
test_data  <- df[-train_index, ]

```

```{r}
# 1. Prepare data for regression

# Create model matrix for glmnet (converts factors to dummy variables)
# Remove the response variable (median_earnings) 
X <- model.matrix(earnings ~ . , data = df)[,-1]
y <- df$earnings

# Split data into training and testing sets
set.seed(123)
train_indices <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_indices, ]
y_train <- y[train_indices]
X_test <- X[-train_indices, ]
y_test <- y[-train_indices]


```

##### 1. Model without controls

```{r}
model1 <- lm(earnings ~ delivery_mode, data = train_data)
model1_summary <- summary(model1)
model1_summary

# Predict values
model1_predictions <- predict(model1, newdata = test_data)

# Calculate RMSE
model1_rmse <- sqrt(mean((test_data$earnings - model1_predictions)^2))
cat("model1 RMSE:", model1_rmse, "\n")
```

```{r}
# Extract R-squared and Adjusted R-squared
r_squared_m1 <- model1_summary$r.squared
adj_r_squared_m1<- model1_summary$adj.r.squared

# Print values
cat("R-squared:", r_squared_m1, "\n")
cat("Adjusted R-squared:", adj_r_squared_m1, "\n")

```

Based on the above model, there is a negative relationship between earnings and delivery mode when most of the credentials are conducted onsite. This gives us an early indication that onsite programs have a negative effect on earnings. The r2 an the adjusted r2 values are quite low, need to add in more variables to the model.

##### 2. Model with field-level characteristics

```{r}
model2 <- lm(earnings ~ delivery_mode + course_field, data = train_data)
model2_summary<- summary(model2)
model2_summary

```

```{r}
# Extract R-squared and Adjusted R-squared
r_squared_m2 <- model2_summary$r.squared
adj_r_squared_m2 <- model2_summary$adj.r.squared

# Print values
cat("R-squared:", r_squared_m2, "\n")
cat("Adjusted R-squared:", adj_r_squared_m2, "\n")

```

Even after removing non-stem fields, there were 152 unique fields in the dataset. Adding the field level characteristics brought down the negative coefficient size of the on-site delivery mode variable and only few feilds, such as Optometry and Nursing, had significant coefficients. The r2 and the adjusted r2 values improved from the previous model however, it can improve further.

##### 3. Model with field level + institutional characteristics

```{r}
model3 <- lm(earnings ~ ., data = train_data)
model3_summary <-summary(model3)
model3_summary
```

```{r}
# Extract R-squared and Adjusted R-squared
r_squared_m3 <- model3_summary$r.squared
adj_r_squared_m3 <- model3_summary$adj.r.squared

# Print values
cat("R-squared:", r_squared_m3, "\n")
cat("Adjusted R-squared:", adj_r_squared_m3, "\n")
```

Significant Variables:

-   Credit level - Undergraduate level is highly significant with a positive effect on earnings while some credit levels, such as first professional degree and graduate/ professional certificate, has a negetive relationship with earnings.

-   School Type- Public and private for profits schools have a negative relationship with earnings

-   No. of Pell Grants - This variable has a positive relationship with earnings.

**Delivery_mode: All credentials online** - The coefficient of this variable is not significant in this model; however it seems to have a small negative effect on earnings.

##### 4. Regularization

1.  Ridge Regression

```{r}
# 1. Prepare data for regularized regression

# Create model matrix for glmnet (converts factors to dummy variables)
# Remove the response variable (median_earnings) 
X <- model.matrix(earnings ~ . , data = df)[,-1]
y <- df$earnings

# Split data into training and testing sets
set.seed(123)
train_indices <- sample(1:nrow(X), 0.8 * nrow(X))
X_train <- X[train_indices, ]
y_train <- y[train_indices]
X_test <- X[-train_indices, ]
y_test <- y[-train_indices]

# Ridge Regression

# Finding optimal lambda using cross-validation
set.seed(123)
ridge_cv <- cv.glmnet(X_train, y_train, alpha = 0, nfolds = 10)
plot(ridge_cv)

# Best lambda value
best_lambda_ridge <- ridge_cv$lambda.min
cat("Best lambda for ridge regression:", best_lambda_ridge, "\n")

# Fit ridge model with optimal lambda
ridge_model <- glmnet(X_train, y_train, alpha = 0, lambda = best_lambda_ridge)
print(ridge_model)

# Get coefficients
ridge_coefs <- coef(ridge_model)
ridge_coefs_df <- data.frame(
  feature = rownames(ridge_coefs),
  coefficient = as.vector(ridge_coefs)
) %>%
  filter(coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

print(ridge_coefs_df)

# Visualize top coefficients
ridge_coefs_df %>%
  filter(feature != "(Intercept)") %>%
  top_n(15, abs(coefficient)) %>%
  ggplot(aes(x = reorder(feature, abs(coefficient)), y = coefficient)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Ridge Regression Coefficients",
       x = "Feature",
       y = "Coefficient Value") +
  theme_minimal()

# Evaluate ridge model on test data
ridge_predictions <- predict(ridge_model, newx = X_test)
ridge_rmse <- sqrt(mean((y_test - ridge_predictions)^2))
cat("Ridge Regression RMSE:", ridge_rmse, "\n")


```

-   Used the optimal lambda value after cross-validation.

-   The plot above shows the top coefficients.

```         
delivery_modeMost credentials onsite                                                           -1.374220e+02
```

The ridge model also gives you a negative coefficient for credentials completed onsite.

```         
```

2.  Lasso Regression

```{r}
# Find optimal lambda using cross-validation
set.seed(123)
lasso_cv <- cv.glmnet(X_train, y_train, alpha = 1, nfolds = 10)
plot(lasso_cv)

# Best lambda value
best_lambda_lasso <- lasso_cv$lambda.min
cat("Best lambda for lasso regression:", best_lambda_lasso, "\n")

# Fit lasso model with optimal lambda
lasso_model <- glmnet(X_train, y_train, alpha = 1, lambda = best_lambda_lasso)
print(lasso_model)

# Get coefficients - lasso will zero out some coefficients
lasso_coefs <- coef(lasso_model)
lasso_coefs_df <- data.frame(
  feature = rownames(lasso_coefs),
  coefficient = as.vector(lasso_coefs)
) %>%
  filter(coefficient != 0) %>%
  arrange(desc(abs(coefficient)))

print(lasso_coefs_df)

# Visualize non-zero coefficients
lasso_coefs_df %>%
  filter(feature != "(Intercept)") %>%
  ggplot(aes(x = reorder(feature, abs(coefficient)), y = coefficient)) +
  geom_col() +
  coord_flip() +
  labs(title = "Non-Zero Lasso Regression Coefficients",
       subtitle = "Variables selected by lasso as important predictors",
       x = "Feature",
       y = "Coefficient Value") +
  theme_minimal()

# Evaluate lasso model on test data
lasso_predictions <- predict(lasso_model, newx = X_test)
lasso_rmse <- sqrt(mean((y_test - lasso_predictions)^2))
cat("Lasso Regression RMSE:", lasso_rmse, "\n")
```

```{r}
# Count non-zero coefficients (excluding intercept)
n_nonzero <- sum(lasso_coefs[-1,] != 0)
n_total <- length(lasso_coefs) - 1  # Subtract 1 for intercept
n_dropped <- n_total - n_nonzero

cat("Total variables:", n_total, "\n")
cat("Variables retained:", n_nonzero, "\n")
cat("Variables dropped:", n_dropped, "\n")
cat("Percentage dropped:", round(n_dropped/n_total*100, 1), "%\n")

```

-   Used the optimal lambda value after cross-validation.

-   The lasso model dropped 40.7% of the variables from the model. According to the model, some of the field-level variables were the top contributors to earnings.

## 3. Reliability and Model Evaluation

```{r}
model_comparison <- data.frame(
  model = c("Ridge", "Lasso", "model1","model2","model3"),
  rmse = c(ridge_rmse, lasso_rmse, model1_rmse, NA, NA),
  r2 = c(NA,NA , r_squared_m1, r_squared_m2, r_squared_m3),
  Adj_r2 = c(NA, NA, adj_r_squared_m1, adj_r_squared_m2, adj_r_squared_m3)
)
model_comparison

library(knitr)
library(kableExtra)

kable(model_comparison, digits = 3, caption = "Model Performance Comparison") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

-   The above table summarizes the model performance across the different models, comparing the RMSE values of the models its clear that the lasso model has the smallest error and based on the models built it best predicts the revenue.

-   In terms of the smart question, **Do online (distance education) programs yield significantly different earnings compared to in-person programs?** several models initially revealed a surprising negative relationship between in-person learning and earnings. However, as additional control variables were introduced, this relationship became less statistically significant, suggesting that the initial effect may have been confounded by other factors.

```         


```
