---
title: "DATS 6101 Midterm Project"
author: "Venkatesh Nagarjuna, Diluni Peiris, Manoj Kodihalli Venkatesh, Jose Luis Garcia Montero, Ella Brown"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options: 
  markdown: 
    wrap: 72
---

```{r init, include=FALSE}
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 5) 
```

# Mid Term Paper

## 1. What does the dataset reveal? - Ella

The dataset did not reveal much because it was artificially generated data from Kaggle. When first looked at, the data appeared to have a good spread of variables, both categorical and numerical. However, after conducting extensive exploratory data analysis, it was revealed that the data was essentially flat. All of the different variables appeared to be very uniformly spread, except for a few. For the most part, the different categories of variables were almost entirely equal, making all of the statistical analysis conducted insignificant. As a result of this, the dataset unfortunately did not reveal much in terms of its values. However, performing the analysis and going through the motion of the code could help for future analysis with other data. 

This dataset contains **10,000+ records** of workout and fitness-related
data collected from various fitness apps and devices. Contains below
features.

**Column Descriptions**

-   **User ID**: A unique numerical identifier assigned to each user in
    the dataset.

-   **Age**: The age of the user, ranging from 18 to 60 years.

-   **Gender**: The gender of the user (Male, Female, or Other).

-   **Height (cm)**: The height of the user in centimeters.

-   **Weight (kg)**: The weight of the user in kilograms.

-   **Workout Type**: The type of workout performed (e.g., Cardio,
    Strength, Yoga, HIIT, Cycling, Running).

-   **Workout Duration (mins)**: The total duration of the workout in
    minutes.

-   **Calories Burned**: The number of calories burned during the
    workout session.

-   **Heart Rate (bpm)**: The average heart rate (beats per minute)
    recorded during the workout.

-   **Steps Taken**: The total number of steps taken during the workout
    (applicable for walking/running workouts).

    Source:
    <https://www.kaggle.com/datasets/adilshamim8/workout-and-fitness-tracker-data>

## 2. Dataset limitations - Ella

The dataset appeared to be artificially generated, as revealed through extensive Exploratory Data Analysis (EDA), which showed it to be flat and evenly distributed. This lack of variability led to minimal or no meaningful correlations among key workout metrics, resulting in statistically insignificant findings. Consequently, any insights drawn from this data are unlikely to generalize to real-world workout patterns, limiting its value for data-driven decision-making. 

## 3. Data collection methods - Ella

The data was taken from the website Kaggle. It is titled “Workout & Fitness Tracker Dataset” and has over 10,000 records of workout and fitness-related data collected from various fitness apps and devices. As previously mentioned, this data was found to be artificially generated, not collected from actual people’s data.

## 4. Previous analyses on related topics - Diluni

Workout data analysis has grown increasingly popular with the rising
accessibility of fitness trackers, smartwatches, and mobile apps that
capture detailed metrics. As more people prioritize an active lifestyle,
particularly the younger generation, the demand for data-driven insights
has surged. Additionally, with health concerns affecting individuals
across all age groups, there is heightened interest in using fitness
data to monitor and improve overall well-being. This growing trend has
fueled extensive research in the field, exploring how workout data can
enhance performance, personalized fitness plans, and even predict
potential health risks.

Bonomi and Westerterp (2012) explored the use of machine learning (ML)
in classifying physical activity types, such as walking, running, and
cycling, based on accelerometer data. Their study showed that ML
algorithms can accurately predict energy expenditure by analyzing
patterns in heart rate, step count, and movement.

Mooney et al. (2018) examined the use of wearable devices and machine
learning models to create personalized fitness plans. Their study showed
that by classifying exercise types and analyzing intensity levels, ML
models could recommend individualized workout adjustments.

Köller et al. (2022) this study explores how sleep data combined with
workout metrics (intensity, duration) can predict recovery status in
endurance athletes. It highlights the importance of multi-modal data
analysis in optimizing performance.

A study was conducted by Amirreza et al. (2020) to evaluate the relation
of sleep duration and quality with blood pressure (BP) and heart rate
variability (HRV). It was found that lower the quality of sleep higher
would be the resting heart rate. 

## 5. How gathered research contributed to question development - Diluni

The research gathered helped us to understand the types of analysis and
techniques we can use to analyse fitness data.

Based on the research conducted by Bonomi and Westerterp (2012) on how
different exercise can have an impact on factors such as heart rate,
calorie burn we formulated a SMART question to understand what
combination of workout types and durations maximizes calorie burn
efficiency for different user profiles. 

Personalized fitness plans was an interesting idea that came up in our
discussions and the research conducted by Mooney et al. (2018) gave us a
guide on how we could explore that possibility using our data. We
created a clustering model to tackle that objective.

We wanted to explore how the BMI category would have an impact on
Calorie burn based on some studies that were done indicating that it
could have distinct effects based on the BMI category. We formulated a
question to explore these effects.

Based on the study that was conducted by Amirreza et al. (2020), this
was an interesting relationship we wanted to explore in our data. If you
do not have enough quality sleep does that impact your resting heart
rate? The research highlighted that it has a negative relationship,
therefore we tried to prove this relationship using our data.

## 6. Additional information that could enhance the analysis - Luis
Despite our attempts to create derived variables like BMI and body type classifications, the analysis remained limited. Additional information needed would include:

Having already exhausted basic derived metrics (BMI, body type) that showed uniform distributions, we need:
  True physiological relationships and correlations
  Non-uniform, realistic distributions of derived variables

Additional real-world variables such as:
  Real-world demographic distributions for gym-goers or fitness enthusiasts
  Actual physiological relationships between variables (e.g., how heart rate correlates with workout intensity and calories burned)
  Time-series data for individual users to track progress
  Seasonal patterns in workout behaviors
  Detailed workout intensity levels
  Diet information beyond daily calorie intake
  Rest periods between workouts
  Previous fitness experience indicators
  Environmental factors (indoor vs outdoor workouts, temperature, humidity)
  Sleep patterns and their impact on workout effectiveness
  Injury history or physical limitations

However, this defeats the purpose of the choice of this dataset in particular, as one of the key reasons was it's ease of use. 

## 7. Changes to your question after EDA - Luis

The initial research questions need to be completely reformulated due to the following discoveries:

  The dataset appears to be simulated rather than real-world data
  Key variables show unrealistic uniform distributions
  There are no meaningful correlations between variables that should naturally correlate
  Demographic distributions (especially gender) contradict known population statistics
  BMI and other derived variables show unnaturally even distributions

Proposed changes:

  First acquire a verified, real-world fitness dataset
  Revise research questions to focus on actual patterns in authentic fitness data
  Include more complex variable relationships that reflect true physiological and behavioral patterns
  Develop questions that can lead to actionable insights for real fitness applications
  Consider narrowing the scope to specific demographic groups or workout types to ensure data quality and relevance

This response reflects the need to pivot from the current approach due to the fundamental data quality issues discovered during EDA.


## 8. Preliminary insights based on EDA - V

Data Structure Issues:

  Dataset revealed to be artificially structured/simulated
  Variables showed unrealistic uniform distributions
  No meaningful correlations between variables that should naturally correlate

Key Distribution Findings:
  Gender was evenly distributed between 'male', 'female', and 'other' (contradicting real-world demographics)
  Uniform distributions observed in:
      Age
      Weight
      Height
      Workout duration
      Daily calorie intake
      Heart rate
      Distance

Partition Analysis Results:
  No significant mean differences when analyzing:
      Burned calories across workout types
      BMI distribution across genders
  Minimal gender differences in most variables
      Only notable difference was in calories burned:
        Females: 546.79
        Males: 557.94
      Difference too small to be practically significant

Derived Variables:
  Created BMI and body type classifications
  These new variables also showed uniform distributions
  No meaningful patterns emerged even with derived metrics

Statistical Analysis Attempts:
  No significant correlations found between any variables
  Regression models yielded:
     Non-significant coefficients
     Coefficients close to zero
     No predictive power

Overall conclusion: The data appears to be synthetic/simulated rather than real-world data, making it unsuitable for meaningful statistical analysis.


## 9. References (APA style preferred) - Manoj

Loading Libraries

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(gridExtra)
library(corrplot)
library(tidyverse)
library(caret)
library(readr)
library(cluster)
library(factoextra)
```

Reading the Data

```{r}
fitnessdf <- data.frame(read.csv("workout_fitness_tracker_data.csv"))
```

# **Data Pre-processing**

-   Removing Redundant Columns

-   Renaming Columns

-   Updating Data Types

-   Handling Missing Values

-   Handling Outliers

-   Feature Creation

**Removing Redundant Columns**

-   The following columns will be removed from the dataset. User ID:
    Unique entries Water Intake (liters), VO2 Max and Body Fat (%):
    Redundant data (same value repeated)

```{r include=FALSE}
fitnessdf_cleaned <- fitnessdf[, !names(fitnessdf) %in% c("User.ID", "Water.Intake..liters.", "VO2.Max", "Body.Fat....")] 
```

```{r}
str(fitnessdf_cleaned)
```

**Renaming Columns**

```{r include=FALSE}
colnames(fitnessdf_cleaned)<- c("Age", "Gender","Height_cm","Weight_kg","Workout_Type","Workout_Duration_mins","Calories_Burned",                   "Heart_Rate_bpm","Steps_Taken","Distance_km","Workout_Intensity","Sleep_Hours","Daily_Calories_Intake",      "Resting_Heart_Rate_bpm","Mood_Before_Workout","Mood_After_Workout")
```

**Updating Data Types**

```{r include=FALSE}
factor_cols <- c("Gender", 
                 "Workout_Type", 
                 "Workout_Intensity", 
                 "Mood_Before_Workout", 
                 "Mood_After_Workout")

fitnessdf_cleaned[factor_cols] <- lapply(fitnessdf_cleaned[factor_cols], function(x) as.factor(x))
str(fitnessdf_cleaned)
```

**Imputation**

```{r}
missing_report <- data.frame(
  Column = names(fitnessdf_cleaned),
  Missing_Count = colSums(is.na(fitnessdf_cleaned)),
  Missing_Percentage = colSums(is.na(fitnessdf_cleaned)) / nrow(fitnessdf_cleaned) * 100
  )

print(missing_report) 
```

\*There are no missing values in the dataset.

**Handling Outliers**

```{r}
# Select numeric columns
fitnessdf_numeric <- fitnessdf_cleaned %>% select(where(is.numeric))

# Convert to long format
fitnessdf_long <- fitnessdf_numeric %>% pivot_longer(
  cols = everything(),
  names_to = "Variable",
  values_to = "Value"
)

# Create box plots in a grid layout
ggplot(fitnessdf_long, aes(x = Variable, y = Value, fill = Variable)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 16, outlier.size = 3) +
  theme_minimal() +
  labs(title = "Boxplots of Numeric Columns") +
  theme(legend.position = "none") +
  facet_wrap(~Variable, scales = "free",ncol= 3)  # Creates a grid of individual plots
```

Detecting outliers using Z-score.

```{r}
# Detecting outliers by Z-score
zscore_outliers <- function(df, threshold = 3) {
  outliers <- data.frame(Row = integer(), Column = character(), Value = numeric())
  
  for (col in colnames(df)) {
    if (is.numeric(df[[col]])) {
      z_scores <- scale(df[[col]])
      rows <- which(abs(z_scores) > threshold)
      
      if (length(rows) > 0) {
        for (r in rows) {
          outliers <- rbind(outliers, data.frame(Row = r, Column = col, Value = df[r, col]))
        }
      }
    }
  }
  
  return(outliers)
}

# Detect outliers using Z-score
outliers_z <- zscore_outliers(fitnessdf_cleaned)
print("Outliers detected using Z-score:")
print(outliers_z)

```

No outliers were observed in the data.

**Feature Creation** BMI: To categorize the dataset based on BMI
categroies as Underweight, Normal, Overweight and Obese BMI Category:
Four classes as Underweight, Normal, Overweight and Obese Age Groups:
Based on the dataset(Ages 18-59), 18-30 years (Youth), 31-45 (Middle
age), 46-59 (Senior)

**BMI**

```{r include=FALSE}
fitnessdf_cleaned$BMI <- fitnessdf_cleaned$Weight_kg / (fitnessdf_cleaned$Height_cm / 100)^2
```

**BMI Category**

Segmenting the data based on BMI

```{r}

fitnessdf_cleaned <- fitnessdf_cleaned %>%
  mutate(BMI_category= case_when(
     BMI < 18.5 ~ "Underweight",
     BMI >= 18.5 & BMI < 24.9 ~ "Normal",
     BMI >= 24.9 & BMI < 29.9 ~ "Overweight",
     BMI >= 29.9 ~ "Obese"

  )) 

fitnessdf_cleaned$BMI_category <- factor(fitnessdf_cleaned$BMI_category, levels = c("Underweight", "Normal", "Overweight","Obese"))
```

**Age Groups**

Age Groups: Based on the dataset(Ages 18-59), 18-30 years (Youth), 31-45
(Middle age), 46-59 (Senior)

```{r, include=T}
fitnessdf_cleaned <- fitnessdf_cleaned %>%
  mutate(Age_group= case_when(
     Age <= 30 ~ "Youth (<= 30 years)",
     Age >= 31 & Age <= 45 ~ "Middle aged (31 to 45 years)",
     Age >= 46 ~ "Seniors (>= 46 years)"
  )) 

fitnessdf_cleaned$Age_group <- factor(fitnessdf_cleaned$Age_group, levels = c("Youth (<= 30 years)", "Middle aged (31 to 45 years)", "Seniors (>= 46 years)"))
```

# **EDA**

### EDA: Demographic Analysis

```{r}
# Summary statistics for demographic variables
demographics_summary <- fitnessdf_cleaned %>% 
  select(Age, Gender, Height_cm, Weight_kg, BMI_category) %>% 
  summary()
print(demographics_summary)

# Age Distribution Histogram
ggplot(fitnessdf_cleaned, aes(x = Age)) + 
  geom_histogram(bins = 10, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(title = "Age Distribution", x = "Age", y = "Count") +
  theme_minimal()
```

```{r}
# Age Category distribution 
ggplot(fitnessdf_cleaned, aes(x = Age_group, fill = Age_group)) +
  geom_bar() +
  labs(title = "Distribution of Age groups", x = "Age groups", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") 
```

```{r}
#BMI- Category Distibution
ggplot(fitnessdf_cleaned, aes(x = BMI_category, fill = BMI_category)) +
  geom_bar() +
  labs(title = "Distribution of BMI Categories", x = "BMI Category", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")  # Adds nice colors

```

```{r}
# Gender Distribution Bar Plot (if the 'gender' column exists)
ggplot(fitnessdf_cleaned, aes(x = Gender)) +
    geom_bar(fill = "coral", color = "black", alpha = 0.7) +
    labs(title = "Gender Distribution", x = "Gender", y = "Count") +
    theme_minimal()

```

```{r}
# Gender Distribution based on BMI Category
ggplot(fitnessdf_cleaned,aes(x=Gender, fill= BMI_category))+
  geom_bar(position = "dodge")+
  labs(title = "Gender Distribution based on BMI Category", x = "Gender", y = "Count")+
  theme_minimal()+
  scale_fill_brewer(palette = "Set2")
  
```

### EDA: Performance Metrics

```{r}
# Summary statistics for performance metrics
performance_summary <- fitnessdf_cleaned %>% 
  select(Workout_Duration_mins, Calories_Burned, Heart_Rate_bpm) %>% 
  summary()
print(performance_summary)

# Scatter Plot: Workout Duration vs. Calorie Burn
if(all(c("Workout_Duration_mins", "Calories_Burned") %in% colnames(fitnessdf_cleaned))){
  ggplot(fitnessdf_cleaned, aes(x = Workout_Duration_mins, y = Calories_Burned)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    labs(title = "Workout Duration vs. Calorie Burn", 
         x = "Workout Duration (min)", y = "Calories Burned") +
    theme_minimal()
}
```

```{r}
# Heart Rate Distribution Histogram
ggplot(fitnessdf_cleaned, aes(x = Heart_Rate_bpm)) +
    geom_histogram(bins = 15, fill = "purple", color = "black", alpha = 0.7) +
    labs(title = "Heart Rate Distribution", 
         x = "Heart Rate (bpm)", y = "Frequency") +
    theme_minimal()

```

```{r}
#Boxplot: Mood before workout and sleep hours 
ggplot(fitnessdf_cleaned, aes(y = Sleep_Hours, x= Mood_Before_Workout)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "#90EE90")+
  labs(title = "Boxplot of Sleep Hours across Different Moods Before Workout", x = "Mood_Before_Workout", y = "Sleep_Hours") +
  theme_minimal() 
```

If you have had more hours of sleep could assume that your mood before
workout would be better.

### EDA: Workout Patterns

```{r}
# Frequency of Workout Types (if the 'workout_type' column exists)
if("Workout_Type" %in% colnames(fitnessdf_cleaned)){
  ggplot(fitnessdf_cleaned, aes(x = Workout_Type)) +
    geom_bar(fill = "orange", color = "black", alpha = 0.7) +
    labs(title = "Frequency of Workout Types", 
         x = "Workout Type", y = "Count") +
    theme_minimal()
  
  # Boxplot: Workout Duration by Workout Type
  ggplot(fitnessdf_cleaned, aes(x = Workout_Type, y = Workout_Duration_mins)) +
    geom_boxplot(fill = "lightblue", alpha = 0.7) +
    labs(title = "Workout Duration by Workout Type", 
         x = "Workout Type", y = "Duration (min)") +
    theme_minimal()
}
```

### EDA: Gender Mean differences

```{r}
fitnessdf_gender <- fitnessdf_cleaned %>%
  filter(Gender %in% c("Male", "Female"))

gender_summaries <- fitnessdf_gender %>%
  group_by(Gender) %>%
  summarise(
    count = n(),
    mean_height = mean(Height_cm),
    sd_height = sd(Height_cm),
    mean_weight = mean(Weight_kg),
    sd_weight = sd(Weight_kg),
    mean_calories = mean(Calories_Burned),
    sd_calories = sd(Calories_Burned),
    mean_duration = mean(Workout_Duration_mins),
    sd_duration = sd(Workout_Duration_mins)
  )

print("Summary Statistics by Gender:")
print(gender_summaries)

# Perform t-tests for various metrics
t_test_results <- list(
  Height = t.test(Height_cm ~ Gender, data = fitnessdf_gender),
  Weight = t.test(Weight_kg ~ Gender, data = fitnessdf_gender),
  Duration = t.test(Workout_Duration_mins ~ Gender, data = fitnessdf_gender),
  Calories = t.test(Calories_Burned ~ Gender, data = fitnessdf_gender),
  Heart_Rate = t.test(Heart_Rate_bpm ~ Gender, data = fitnessdf_gender),
  Steps = t.test(Steps_Taken ~ Gender, data = fitnessdf_gender),
  Distance = t.test(Distance_km ~ Gender, data = fitnessdf_gender),
  Sleep = t.test(Sleep_Hours ~ Gender, data = fitnessdf_gender),
  Daily_Calories = t.test(Daily_Calories_Intake ~ Gender, data = fitnessdf_gender),
  Resting_HR = t.test(Resting_Heart_Rate_bpm ~ Gender, data = fitnessdf_gender)
)

# Create a summary table of t-test results
t_test_summary <- data.frame(
  Variable = names(t_test_results),
  P_Value = sapply(t_test_results, function(x) x$p.value),
  Mean_Difference = sapply(t_test_results, function(x) diff(x$estimate)),
  T_Statistic = sapply(t_test_results, function(x) x$statistic)
)

# Add significance indicators
t_test_summary$Significance <- ifelse(t_test_summary$P_Value < 0.001, "***",
                                    ifelse(t_test_summary$P_Value < 0.01, "**",
                                           ifelse(t_test_summary$P_Value < 0.05, "*", "ns")))

print("T-Test Results Summary:")
print(t_test_summary)
    
```

## Data distributions

```{r}

numeric_cols <- fitnessdf_cleaned %>%
  select("Age", "Height_cm", "Weight_kg",
         "Workout_Duration_mins", "Calories_Burned",
         "Heart_Rate_bpm", "Steps_Taken", "Distance_km",
         "Workout_Intensity", "Sleep_Hours", "Daily_Calories_Intake",     
         "Resting_Heart_Rate_bpm") %>%
  mutate(across(everything(), as.numeric)) %>%  # Convert all columns to numeric
  as.data.frame()

summary_stats <- summary(numeric_cols)
print(summary_stats)

plot_distributions <- function(data) {
  plots <- list()
  
  for (col in names(data)) {
    if (is.numeric(data[[col]])) {  # Ensure the column is numeric
      p <- ggplot(data, aes(x = .data[[col]])) +
        geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black") +
        geom_density(color = "red") +
        ggtitle(col) +
        theme_minimal()
      
      plots[[length(plots) + 1]] <- p
    }
  }
  
  if (length(plots) > 0) {
    do.call(grid.arrange, c(plots, ncol = 3))  # Arrange all plots in a grid with 3 columns
  } else {
    print("No numeric columns to plot.")
  }
}

plot_distributions(numeric_cols)

```

## Population Mean Hipotesis Tests

```{r}
# Perform t-tests for various metrics
metrics <- c("Height_cm", "Weight_kg", "Workout_Duration_mins", "Calories_Burned", 
             "Heart_Rate_bpm", "Steps_Taken", "Distance_km", "Sleep_Hours", 
             "Daily_Calories_Intake", "Resting_Heart_Rate_bpm")

t_test_results <- list()
# you need to specify var.equal = True to force a regular t-test
for (metric in metrics) {
  t_test_results[[metric]] <- t.test(get(metric) ~ Gender, data = fitnessdf_gender, var.equal = TRUE)
  print(paste("\nT-test Results for", metric, ":"))
  print(t_test_results[[metric]])
}

# Create and display boxplots for all tested metrics with larger plot sizes
plot_list <- list()
for (metric in metrics) {
  p <- ggplot(fitnessdf_gender, aes(x = Gender, y = get(metric), fill = Gender)) +
    geom_boxplot() +
    labs(title = paste(metric),
         y = metric) +
    theme_minimal() +
    theme(plot.title = element_text(size = 16, face = "bold"),
          axis.text = element_text(size = 14),
          axis.title = element_text(size = 14))
  
  plot_list[[length(plot_list) + 1]] <- p
}

# Display plots in three groups
grid.arrange(grobs = plot_list[1:4], ncol = 2)
grid.arrange(grobs = plot_list[5:8], ncol = 2)
grid.arrange(grobs = plot_list[9:10], ncol = 2)

```

## Correlation analysis

```{r correlations}

# Calculate correlations
correlations <- cor(numeric_cols)

# View correlations (rounded to 2 decimal places)
round(correlations, 2)

# Create a basic correlation plot
library(corrplot)
corrplot(correlations, method = "color", 
         type = "upper", 
         addCoef.col = "black", # Add correlation coefficients
         number.cex = 0.7,      # Size of correlation coefficients
         tl.col = "black",      # Text label color
         tl.srt = 45)           # Rotate text labels


```

**SMART Q1**

What combination of workout types and durations maximizes calorie burn
efficiency for different user profiles (clustered by age, BMI, and
fitness level)?

```{r}


# Calculate BMI and create age groups (I'm hardcoding some things to fit it from my own file)
fitnessdf_cleaned <- fitnessdf_cleaned %>%
  mutate(
    BMI = Weight_kg / ((Height_cm / 100)^2),
    BMI_Category = case_when(
      BMI < 18.5 ~ "Underweight",
      BMI >= 18.5 & BMI < 25 ~ "Normal",
      BMI >= 25 & BMI < 30 ~ "Overweight",
      BMI >= 30 ~ "Obese"
    ),
    Calorie_Efficiency = Calories_Burned / Workout_Duration_mins,
    Age_group = case_when(
      Age <= 30 ~ "18-30",
      Age > 30 & Age <= 45 ~ "31-45",
      Age > 45 ~ "46+"
    )
  )

# Visualize the data
ggplot(fitnessdf_cleaned, aes(x = Age_group, y = Calorie_Efficiency, fill = BMI_Category)) +
  geom_boxplot() +
  labs(title = "Calorie Efficiency by Age Group and BMI Category",
       x = "Age Group",
       y = "Calorie Efficiency (Calories/Minute)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Linear regression including BMI category
model <- lm(Calorie_Efficiency ~ Workout_Duration_mins + Workout_Type + Age + BMI_Category, 
            data = fitnessdf_cleaned)
print(summary(model))


```

# **SMART Q2**

How does the relationship between heart rate patterns and calorie burn
vary across different workout types, and can this predict optimal
workout intensity?

```{r}
ggplot(fitnessdf_cleaned, aes(x = Heart_Rate_bpm, y = Calories_Burned, color = Workout_Type)) +
  geom_point(alpha=0.6) +
  geom_smooth(method = 'lm') +
  labs(title = "Heart Rate vs. Calories Burned by Workout Type",
       x = "Heart Rate (bpm)",
       y = "Calories Burned")


```

```{r}
# Find optimal intensity based on calories per heart rate
optimal_intensity <- fitnessdf_cleaned %>%
  mutate(Efficiency = Calories_Burned / Heart_Rate_bpm) %>%
  group_by(Workout_Type, Workout_Intensity) %>%
  summarise(Avg_Efficiency =  round(mean(Efficiency, na.rm = TRUE), 2)) %>%
  arrange(Workout_Type, desc(Avg_Efficiency))

# Display optimal intensity results
print(optimal_intensity)

```

```{r}
# Visualization of optimal intensities
ggplot(optimal_intensity, aes(x=Workout_Type, y=Avg_Efficiency, fill=Workout_Intensity)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title="Optimal Workout Intensity by Workout Type",
       x="Workout Type", 
       y="Calories Burned per bpm")
```

```{r}


# Statistical modeling: Create regression models for each Workout_Type
models <- fitnessdf_cleaned %>% 
  group_by(Workout_Type) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(Calories_Burned ~ Heart_Rate_bpm + Workout_Intensity, data = .x)))

# Model summaries
models_summary <- models %>% 
  mutate(tidied = map(model, broom::tidy)) %>% 
  unnest(tidied)

print(models_summary)

# Evaluate Model Performance: R-squared values for each workout type
model_performance <- models %>%
  mutate(R_squared = map_dbl(model, ~summary(.)$r.squared)) %>%
  select(Workout_Type, R_squared)

print(model_performance)

```

Conclusion: No significant effect on workout intensity based on the
heart rate and calories burned according to the dataset.

# **SMART Q3**

Can we identify distinct user profiles based on their workout patterns
using clustering analysis on workout duration, calorie burn, and heart
rate?

```{r}
cluster_data <- fitnessdf_cleaned %>%
  select(Workout_Duration_mins, Calories_Burned, Heart_Rate_bpm) %>%
  drop_na()

# Add index as unique identifier
cluster_data <- cluster_data %>% 
  mutate(Index = row_number())

# Scale the data for clustering
scaled_data <- cluster_data %>%
  select(-Index) %>%
  scale()

# Determine optimal number of clusters using the elbow method
fviz_nbclust(scaled_data, kmeans, method = "wss") + 
  labs(title = "Optimal Number of Clusters (Elbow Method)")

# Determine optimal number of clusters using Silhouette Method
fviz_nbclust(scaled_data, kmeans, method = "silhouette") +
  labs(title = "Optimal Clusters: Silhouette Method")

```

```{r}

# Perform K-means clustering (Choose optimal k from silhouette results, e.g., k=3)
set.seed(123)
optimal_k <- 6  # Adjust based on silhouette/elbow plots
kmeans_result <- kmeans(scaled_data, centers = optimal_k, nstart = 25)

# Add cluster labels to original data
cluster_results <- cluster_data %>%
  mutate(Cluster = factor(kmeans_result$cluster))

# Visualize clusters using PCA
fviz_cluster(kmeans_result, data = scaled_data,
             ellipse.type = "euclid",
             repel = TRUE,
             ggtheme = theme_minimal()) +
  labs(title = "Workout Patterns Clustering")

# Summarize each cluster's characteristics
cluster_summary <- cluster_results %>%
  group_by(Cluster) %>%
  summarise(
    Count = n(),
    Avg_Workout_Duration = round(mean(Workout_Duration_mins), 2),
    Avg_Calories_Burned = round(mean(Calories_Burned), 2),
    Avg_Heart_Rate = round(mean(Heart_Rate_bpm), 2)
  )

# Display summary
print(cluster_summary)

```

Cluster characteristics:

Cluster 1 (Moderate Duration, Low Calorie, High Intensity): - High
intensity, cardio-focused workouts

Cluster 2 (Moderate Duration, High Calorie, Low Intensity): - Very high
calorie burn - Low intensity, strength-focused or resistance workouts

Cluster 3 (Short Duration, High Calorie, High Intensity): - Short but
intense workouts, possibly HIIT

Cluster 4 (Long Duration, High Calorie, High Intensity): - Extended
cardio or endurance sessions

Cluster 5 (Short Duration, Low Calorie, Moderate Intensity): - Short
duration, moderate effort, possibly low-intensity cardio or stretching

Cluster 6 (Long Duration, Moderate Calorie, Low Intensity): - Extended
low-intensity workouts, e.g., walking or yoga

# **SMART Q4**

1\. Effects of workout type on different BMI categories [Factorial
ANOVA]

Factor 1: BMI Category - Underweight, Normal, Overweight and Obese

Factor 2 : Workout Type - Cycling, Cardio, HIIT, Strength, Yoga, Running

Dependent Variable: Calories_Burned

[Main Effects]{.underline}

BMI Category:

$$
H_0: \mu_{underweight} = \mu_{normal} = \mu_{overweight} = \mu_{obese}
$$\
$$
H_1: \text{ At least one mean is different}
$$

The calorie burn is the same for all BMI levels.

```{r}
#Boxplot: Calories_Burned across different BMI levels 
ggplot(fitnessdf_cleaned, aes(y = Calories_Burned, x= BMI_category)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "#90EE90")+
  labs(title = "Boxplot of Calories_Burned across different BMI Categories", x = "BMI_category", y = "Calories_Burned") +
  theme_minimal() 
```

Workout Type:

$$
H_0: \mu_{cycling} = \mu_{Cardio} = \mu_{HIIT} = \mu_{Strengh} = \mu_{yoga} = \mu_{running} 
$$

$$
H_1: \text{ At least one mean is different}
$$

The calorie burn is the same for all workout types.

```{r}
#Boxplot: Boxplot ofCalories_Burned across different Workout types 
ggplot(fitnessdf_cleaned, aes(y = Calories_Burned, x= Workout_Type)) +
  geom_boxplot(outlier.colour = "red", outlier.shape=16, outlier.size=4, fill = "lightblue")+
  labs(title = "Boxplot of Calories_Burned across different Workout types", x = "Workout_type", y = "Calories_Burned") +
  theme_minimal() 
```

[Interaction Effect]{.underline}

$$
H_0:  \text{There is no interaction between BMI level and workout type on Calorie burn}
$$

$$
H_1:  \text{There is an interaction between BMI level and workout type on Calorie burn}
$$

The workout type does not have an influence on Calorie burn based on BMI
Levels.

```{r, include=T}
# Run factorial ANOVA
result <- aov(Calories_Burned ~ BMI_category * Workout_Type, data = fitnessdf_cleaned)
```

```{r, include=T}
# Run factorial ANOVA
summary(result)
```

Based on the Results;(Testing at 95% significance level)

1\. Main effects

-   BMI Level: Cannot Reject HO as P value is \> 0.05,The average
    Calorie burn is equal among different BMI levels.
-   Workout Type: Cannot Reject HO as P value is \> 0.05, The average
    Calorie burn is equal among different Workout types.

2.  Interaction Effect

-   Cannot Reject HO as P value is \> 0.05, Different BMI groups respond
    the same to different workout types.

# **SMART Q5**

Link between poor sleep and resting heart rate (fewer the hours of sleep
and higher resting heart rate) Source:
<https://pmc.ncbi.nlm.nih.gov/articles/PMC7680518/>

```{r, include=T}
# Scatter Plot
ggplot(fitnessdf_cleaned, aes(x = Sleep_Hours, y = Resting_Heart_Rate_bpm)) +
    geom_point(alpha = 0.6, color = "darkgreen") +
    labs(title = "Sleep Hours vs. Resting Heart Rate (bpm)", 
         x = "Sleep_Hours", y = "Resting Heart Rate (bpm)") +
    theme_minimal()
```

```{r, include=T}
cor(fitnessdf_cleaned$Sleep_Hours,fitnessdf_cleaned$Resting_Heart_Rate_bpm)
```

Coefficient - -0.0125. It is negatively correlated as expected, however
the strength of the relationship is low.

**SMART Q6** Is there a relationship between mood before and mood after
workout?

Mood Before Workout: Happy, Neutral, Stressed and Tired.

Mood After Workout: Energized, Fatigued and Neutral

Using a Chi-squared test of Independence.

$$
H_0: \text{There is no relationship between mood before and after workout} 
$$

$$
H_1: \text{There is a relationship between mood before and after workout} 
$$

```{r, include=T}
# Creating a Contingency Table
table_data <- table(fitnessdf_cleaned$Mood_Before_Workout, fitnessdf_cleaned$Mood_After_Workout)
print(table_data)
```

```{r, include=T}
# Run Chi-Square Test
chi_sq_result <- chisq.test(table_data)
print(chi_sq_result)
```

P-value is \> 0.05, Therefore we cannot reject the Null. There's no
relationship between mood before and after workout.
